Precise Spiking Motifs in Neurobiological and Neuromorphic~Data

### 2022-12-28 brainsci-2069565_accepted.tex
MDPI: Please change the (a) into (d) in picture. Fixed !

That results suggest that precise spiking motifs are not necessarily grouped on a topography and can be forming apparently randomly arranged connections. Interestingly, one theoretical viewpoint considers synfire braids~\citep{bienenstock_model_1995}, where a precise sequential motif of spikes will synchronize as it reaches the soma of a neuron for which synaptic delays are adequately tuned. Furthermore, computational modeling shows that at the scale of neurons, an efficient neural code can emerge where spike times are organized in prototypical, precise temporal motifs~\citep{izhikevich_polychronization_2006}, which this author defined as \emph{\hl{polychronous groups}}. The rest of this review will be devoted to present evidence for the use of such precise spiking motifs in computational neuroscience, neurobiology and neuromorphic engineering. As a summary, it seems that such precise structural information is essential to the neural code and that it seems imperative to include this information in the decoding algorithm for a better understanding of neural activity.

@@

In this section, we will review current evidence on how we may take advantage of spiking motifs in neurobiology, that is, in an effort to understand actual recordings from biological neural tissues. In most generic computational models, the neural activity is assumed to be encoded in the firing rate. For instance, the output of the so-called linear non-linear (L-NL) models is assumed to model the response of a biological neuron as the sequence of a linear integration followed by a non-linear spiking response generating spikes according to a Poisson point process\note[Laurent]{adding reference}~\citep{simoncelli_characterization_2003}. As such, a simple decoding strategy is to infer the input knowing the neuron's tuning curves, that is, its selectivity to a range of features~\citep{jazayeri_optimal_2006} or simply by a simple regression~\citep{berens_fast_2012}. This latter model assumes a Bernoulli model for the generation of spikes, such that the decoding amounts to a single-layer logistic regression. An important perspective of these methods used to decipher the recorded activity is that it could be ultimately used to fit neural network models to the recorded activity~\citep{bellec_fitting_2021}. In this particular paper, the authors fit the summary statistics of neural data with a (differentiable) spiking network model. The loss function is the cross entropy (following the Bernouilli hypothesis with a GLM, where each unit is modelled with an SRM neuron~\citep{gerstner_time_1995}) and embedded with recurrent dynamics. In particular, it comes with \href{https://github.com/EPFL-LCN/pub-bellec-wang-2021-sample-and-measure}{\hl{code}} and uses the \note[Laurent]{now citing the URL}publicly available V1 dataset~\citep{kohn_utah_2016}, which allows supervising the model, with the input being the movie and the output the spikes recorded. These type of model may infer sparse activity in a set of binary latent variables, each describing the activity of a cell assembly~\citep{warner_probabilistic_2022}.

@@

\begin{figure}[H]
%\centering
\includegraphics[width=\linewidth]{figures/Diesmann_et_al_1999_spade.pdf}
\caption{
  Detecting motifs using SPADE. 
  We used the SPADE algorithm~\citep{stella_3d-spade_2019} by adapting their \href{https://elephant.readthedocs.io/en/latest/tutorials/spade.html}{\hl{tutorial}} on the data generated in  Figure~\ref{fig:diesman}. This allowed to label different precise spike motifs which are denoted by different colors. Spikes belonging to the same motif have the same color. %As the synfire chain gets more precise, the detected patterns grow in size.
}\label{fig:Diesmann_et_al_1999_spade}
\end{figure}

@@

However, it was not until the study of the interaural time difference (ITD) by \remove[Amélie]{gerstnerneuronal1996}~\citet{carr_circuit_1990} that it was discovered that this delay is not homogeneous for all neurons of the same type and species, but adapts according to their function. This ITD is a biological mechanism which allows for the azimuthal localization of sound by barn owls, by organically computing the difference in arrival time of a sound between their two ears. \add[Amélie]{It was first theorized in the Jeffress ``ITD-versus-place model''}~\citep{jeffress_place_1948}. \add[Amélie]{As hypothesized by this model, the \textit{\hl{nucleus laminaris}} of the avian brain contains coincidence detectors and, associated with the \textit{\hl{nucleus magnocellularis}} axons, forms circuits for processing ITD}~\citep{konishi_coding_2003}. According to~\citep{gerstner_neuronal_1996}, there is a true paradox in auditory neural systems since ``neural networks encode behaviorally relevant signals in the range of a few $\mu s$ with neurons that are at least one order of magnitude slower'' and therefore necessarily need to play on the response time to do so. \add[Amélie]{This assertion confused the mean interspike interval, i.e., how often a neuron can fire, and the specific spike time, i.e., how precisely a spike can be generated. However, it has nevertheless contributed to the recognition of the importance of time in various biological mechanisms.} \remove[Amélie]{penacochlear2001 conducted an extended study on ITD in order to confirm this theory by identifying the coincidence detector involved, which the authors situated in the nuclear laminaris. In the mouse somatosensory cortex, camontiming2019 found a strong correlation between the delay of the mouse behavioral response and the timing of multiunit activity evoked by a trained whisker.} A first hypothesis suggested that the sound coincidence was detected using stereausis, i.e., the temporal disparity between the left and right cochlear loci in the owl's brain. However, it was quickly set aside, as the predictions did not match the measured disparities in the loci, and no variation was perceived in the nuclear laminaris for a similar sound intensity in both owl ears. The authors supported a second hypothesis, that of different axonal delays in the ipsi- and contra-lateral cochlear nucleus magnocellularis.~\citet{seidl_mechanisms_2010} experimentally seconds this hypothesis of a ``coarse'' regulation of delay, as the authors concluded that regulations at different sites within individual axons of at least two parameters, namely, the axon diameter and internode distances, might be responsible for precise adjustments of physiological delays, thus allowing the ITD detection. \add[Amélie]{The authors also noted that the barn owl's axons seem to change in length, thus implementing a ``pure delay line''.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Interplay of Delay Adaptation and Neural Activity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% How is the neuronal activity detected ?
However, one question remains: how do oligodendrocytes detect neuronal activity and regulate the myelinization accordingly? To answer this, we must first study the myelinization process. The OPCs first proliferate in the white matter \textit{via} a self-repulsive process, thus allowing for an evenly spaced network, and identify target axons. Most OPCs then differentiate into oligodendrocytes and immediately initiate myelinization, with no further migration~\citep{simons_oligodendrocytes_2016}. Not much is known about how oligodendrocytes select the axons to myelinate, but it seems that myelination only takes place on large enough axons and is strongly regulated by several factors~\citep{kuhn_oligodendrocytes_2019}, such as Ca$^{2+}$ activity~\citep{baraban_ca2_2018} of the neuregulin 1 growth factor~\citep{nave_axonal_2006}. The important role of myelination on delay learning and biological behaviors, as highlighted in the previous paragraph, suggests that the identification of target neurons as well as the myelin production is also regulated by neuronal activity. Indeed,~\citet{cullen_periaxonal_2021} experimentally demonstrated that learning and associated neuronal activity modify the Ranvier nodes' length and the periaxonal space width in the adult mouse brain. They also confirmed that the delay correlates with the level of skill acquisition.~\citet{gibson_neuronal_2014} suggests that neuronal activity does not solely promote adaptive myelination in the mammalian brain, but also OPC differentiation and oligodendrogenesis. Some further studies show that oligodendrocytes may detect neuronal activity thanks to growth factors or neurotransmitters released through ion channels or {\hl{via}} exocytosis, but does not require any axo-glial synaptic communication~\citep{fields_new_2015}.

@@

\subsection{Izhikevich's Polychronization Model}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% polychronization
As we saw, most SNN, and in particular those adapted from analogous deep-learning-like architectures, rely on an encoding of information based on a continuously varying firing rate. Notable exceptions of SNNs using precise spike timings are the \add[Antoine]{time-encoding machine by}~\citet{lazar_time_2004} and \textit{\hl{polychronization}} model of~\citet{izhikevich_polychronization_2006}. \add[Antoine]{In this section, we focus on the polychronization model} based on a random recurrent model of spiking neurons, including synaptic delays chosen from a range of biologically realistic delays (from $0$ to $20~\ms$) and whose weights evolved with a spike-time-dependent plasticity (STDP) learning rule~\citep{markram_regulation_1997}. \change[Antoine]{It is consensus that spike timing (STDP) plays a crucial role in the development of synaptic efficacy for many different kinds of neurons}{It was shown that spike timing (STDP) has an impact on the development of synaptic efficacy for many kinds of neurons}~\citep{caporale_spike_2008}. Delays are defined as the total time taken for a spike to be conducted from one presynaptic neuron's soma to the efferent postsynaptic neuron's soma. It is worth mentioning that only the weights are changed using the STDP rule and that the set of delays is set randomly at initialization and that delays are then ``frozen'' for the rest of the simulation. Due to the interplay between the delays and STDP, the spiking neurons spontaneously self-organize into groups and generate patterns of stereotypical polychronous activity, i.e.,~exhibit reproducible time-locked firing patterns which  the author defined as ``polychronous groups'' (PGs). One core ingredient of this model is the fact that the neurons composing a group fire at different times, but due to the heterogeneous delays, the spikes reach the postsynaptic neuron at the same time. This synchrony of arrival at the soma of the neuron leads to the summation of the excitatory post-synaptic potentials evoked by each spike, and thus to the crossing of the voltage threshold and to the discharge of a spike (see Figure~\ref{fig:izhikevich}). According to the STDP rule, the group of neurons involved in this polychronous activity will see their synaptic weight increase and, thus, may consolidate the formation of a polychronous group. 
%----------------------------%
%

@@

\begin{figure}[H]%[t!][]
%  \centering
  \includegraphics[width=0.490\linewidth]{figures/THC_1a_k.pdf}
  \includegraphics[width=0.490\linewidth]{figures/THC_1c.pdf}
  \includegraphics[width=0.490\linewidth]{figures/THC_1b.pdf}
  \includegraphics[width=0.490\linewidth]{figures/THC_1a.pdf}
    \caption{
      Detecting event-based motifs using spiking neurons with heterogeneous delays.
    { (\textbf{a})}~Given a generic raster plot defined by a set of spikes occurring on specific addresses and at specific times, one may consider that this information consists of the repeated occurrence of groups of precise motifs of spikes that define ``polychronous groups'' (PGs). 
    { (\textbf{b})}~The generative model is defined by this set of motifs (here $4$ PGs) each defined by  synapses defined by weights (increasing with the radius of the black dots) with heterogeneous delay. 
    { (\textbf{c})}~Generalizing the core polychrony detection model (see Figure~\ref{fig:izhikevich}), one can define a layer of neurons that detect the identity and timing of these PGs~\citep{grimaldi_learning_2022}. % such that the addresses are the synapses of these neurons (which we therefore call here synapses).
    %Note that each afferent may be connected with multiple weights at different delays. The propagation of the afferent information through these delay may generate at each time step a synchronous motif on a subset of synapses. %These PG motifs were efficiently learned by a logistic regression.
    %{\bf (c)}~The output of the model provides with the predicted probability of occurrence of each PG motif at any time, which may be used to generate a spike as a Bernoulli trial, providing in this particular case with an exact identification of PGs occurrences.
    { (\textbf{d})}~\hl{Knowing the} %MDPI: Please change the (a) into (d) in picture.
 results of this detection, one may for illustration purposes highlight them by different colors in the raster plots, showing that in this synthetic example, all spikes are now associated with a PG. 
     }
  \label{fig:THC} % TODO: citer biol cyb fig:THC
\end{figure}

>>>>

That results suggest that precise spiking motifs are not necessarily grouped on a topography and can be forming apparently randomly arranged connections. Interestingly, one theoretical viewpoint considers synfire braids~\citep{bienenstock_model_1995}, where a precise sequential motif of spikes will synchronize as it reaches the soma of a neuron for which synaptic delays are adequately tuned. Furthermore, computational modeling shows that at the scale of neurons, an efficient neural code can emerge where spike times are organized in prototypical, precise temporal motifs~\citep{izhikevich_polychronization_2006}, which this author defined as polychronous groups. The rest of this review will be devoted to present evidence for the use of such precise spiking motifs in computational neuroscience, neurobiology and neuromorphic engineering. As a summary, it seems that such precise structural information is essential to the neural code and that it seems imperative to include this information in the decoding algorithm for a better understanding of neural activity.

@@

In this section, we will review current evidence on how we may take advantage of spiking motifs in neurobiology, that is, in an effort to understand actual recordings from biological neural tissues. In most generic computational models, the neural activity is assumed to be encoded in the firing rate. For instance, the output of the so-called linear non-linear (L-NL) models is assumed to model the response of a biological neuron as the sequence of a linear integration followed by a non-linear spiking response generating spikes according to a Poisson point process\note[Laurent]{adding reference}~\citep{simoncelli_characterization_2003}. As such, a simple decoding strategy is to infer the input knowing the neuron's tuning curves, that is, its selectivity to a range of features~\citep{jazayeri_optimal_2006} or simply by a simple regression~\citep{berens_fast_2012}. This latter model assumes a Bernoulli model for the generation of spikes, such that the decoding amounts to a single-layer logistic regression. An important perspective of these methods used to decipher the recorded activity is that it could be ultimately used to fit neural network models to the recorded activity~\citep{bellec_fitting_2021}. In this particular paper, the authors fit the summary statistics of neural data with a (differentiable) spiking network model. The loss function is the cross entropy (following the Bernouilli hypothesis with a GLM, where each unit is modelled with an SRM neuron~\citep{gerstner_time_1995}) and embedded with recurrent dynamics. In particular, it comes with \href{https://github.com/EPFL-LCN/pub-bellec-wang-2021-sample-and-measure}{code} and uses the \note[Laurent]{now citing the URL}publicly available V1 dataset~\citep{kohn_utah_2016}, which allows supervising the model, with the input being the movie and the output the spikes recorded. These type of model may infer sparse activity in a set of binary latent variables, each describing the activity of a cell assembly~\citep{warner_probabilistic_2022}.

@@

\begin{figure}[H]
%\centering
\includegraphics[width=\linewidth]{figures/Diesmann_et_al_1999_spade.pdf}
\caption{
  Detecting motifs using SPADE. 
  We used the SPADE algorithm~\citep{stella_3d-spade_2019} by adapting their \href{https://elephant.readthedocs.io/en/latest/tutorials/spade.html}{tutorial} on the data generated in  Figure~\ref{fig:diesman}. This allowed to label different precise spike motifs which are denoted by different colors. Spikes belonging to the same motif have the same color. %As the synfire chain gets more precise, the detected patterns grow in size.
}\label{fig:Diesmann_et_al_1999_spade}
\end{figure}

@@

However, it was not until the study of the interaural time difference (ITD) by \remove[Amélie]{gerstnerneuronal1996}~\citet{carr_circuit_1990} that it was discovered that this delay is not homogeneous for all neurons of the same type and species, but adapts according to their function. This ITD is a biological mechanism which allows for the azimuthal localization of sound by barn owls, by organically computing the difference in arrival time of a sound between their two ears. \add[Amélie]{It was first theorized in the Jeffress ``ITD-versus-place model''}~\citep{jeffress_place_1948}. \add[Amélie]{As hypothesized by this model, the \textit{nucleus laminaris} of the avian brain contains coincidence detectors and, associated with the \textit{nucleus magnocellularis} axons, forms circuits for processing ITD}~\citep{konishi_coding_2003}. According to~\citep{gerstner_neuronal_1996}, there is a true paradox in auditory neural systems since ``neural networks encode behaviorally relevant signals in the range of a few $\mu s$ with neurons that are at least one order of magnitude slower'' and therefore necessarily need to play on the response time to do so. \add[Amélie]{This assertion confused the mean interspike interval, i.e., how often a neuron can fire, and the specific spike time, i.e., how precisely a spike can be generated. However, it has nevertheless contributed to the recognition of the importance of time in various biological mechanisms.} \remove[Amélie]{penacochlear2001 conducted an extended study on ITD in order to confirm this theory by identifying the coincidence detector involved, which the authors situated in the nuclear laminaris. In the mouse somatosensory cortex, camontiming2019 found a strong correlation between the delay of the mouse behavioral response and the timing of multiunit activity evoked by a trained whisker.} A first hypothesis suggested that the sound coincidence was detected using stereausis, i.e., the temporal disparity between the left and right cochlear loci in the owl's brain. However, it was quickly set aside, as the predictions did not match the measured disparities in the loci, and no variation was perceived in the nuclear laminaris for a similar sound intensity in both owl ears. The authors supported a second hypothesis, that of different axonal delays in the ipsi- and contra-lateral cochlear nucleus magnocellularis.~\citet{seidl_mechanisms_2010} experimentally seconds this hypothesis of a ``coarse'' regulation of delay, as the authors concluded that regulations at different sites within individual axons of at least two parameters, namely, the axon diameter and internode distances, might be responsible for precise adjustments of physiological delays, thus allowing the ITD detection. \add[Amélie]{The authors also noted that the barn owl's axons seem to change in length, thus implementing a ``pure delay line''.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Interplay of Delay Adaptation and Neural Activity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% How is the neuronal activity detected ?
However, one question remains: how do oligodendrocytes detect neuronal activity and regulate the myelinization accordingly? To answer this, we must first study the myelinization process. The OPCs first proliferate in the white matter \textit{via} a self-repulsive process, thus allowing for an evenly spaced network, and identify target axons. Most OPCs then differentiate into oligodendrocytes and immediately initiate myelinization, with no further migration~\citep{simons_oligodendrocytes_2016}. Not much is known about how oligodendrocytes select the axons to myelinate, but it seems that myelination only takes place on large enough axons and is strongly regulated by several factors~\citep{kuhn_oligodendrocytes_2019}, such as Ca$^{2+}$ activity~\citep{baraban_ca2_2018} of the neuregulin 1 growth factor~\citep{nave_axonal_2006}. The important role of myelination on delay learning and biological behaviors, as highlighted in the previous paragraph, suggests that the identification of target neurons as well as the myelin production is also regulated by neuronal activity. Indeed,~\citet{cullen_periaxonal_2021} experimentally demonstrated that learning and associated neuronal activity modify the Ranvier nodes' length and the periaxonal space width in the adult mouse brain. They also confirmed that the delay correlates with the level of skill acquisition.~\citet{gibson_neuronal_2014} suggests that neuronal activity does not solely promote adaptive myelination in the mammalian brain, but also OPC differentiation and oligodendrogenesis. Some further studies show that oligodendrocytes may detect neuronal activity thanks to growth factors or neurotransmitters released through ion channels or {via} exocytosis, but does not require any axo-glial synaptic communication~\citep{fields_new_2015}.

@@

\subsection{Izhikevich's Polychronization Model}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% polychronization
As we saw, most SNN, and in particular those adapted from analogous deep-learning-like architectures, rely on an encoding of information based on a continuously varying firing rate. Notable exceptions of SNNs using precise spike timings are the \add[Antoine]{time-encoding machine by}~\citet{lazar_time_2004} and polychronization model of~\citet{izhikevich_polychronization_2006}. \add[Antoine]{In this section, we focus on the polychronization model} based on a random recurrent model of spiking neurons, including synaptic delays chosen from a range of biologically realistic delays (from $0$ to $20~\ms$) and whose weights evolved with a spike-time-dependent plasticity (STDP) learning rule~\citep{markram_regulation_1997}. \change[Antoine]{It is consensus that spike timing (STDP) plays a crucial role in the development of synaptic efficacy for many different kinds of neurons}{It was shown that spike timing (STDP) has an impact on the development of synaptic efficacy for many kinds of neurons}~\citep{caporale_spike_2008}. Delays are defined as the total time taken for a spike to be conducted from one presynaptic neuron's soma to the efferent postsynaptic neuron's soma. It is worth mentioning that only the weights are changed using the STDP rule and that the set of delays is set randomly at initialization and that delays are then ``frozen'' for the rest of the simulation. Due to the interplay between the delays and STDP, the spiking neurons spontaneously self-organize into groups and generate patterns of stereotypical polychronous activity, i.e.,~exhibit reproducible time-locked firing patterns which  the author defined as ``polychronous groups'' (PGs). One core ingredient of this model is the fact that the neurons composing a group fire at different times, but due to the heterogeneous delays, the spikes reach the postsynaptic neuron at the same time. This synchrony of arrival at the soma of the neuron leads to the summation of the excitatory post-synaptic potentials evoked by each spike, and thus to the crossing of the voltage threshold and to the discharge of a spike (see Figure~\ref{fig:izhikevich}). According to the STDP rule, the group of neurons involved in this polychronous activity will see their synaptic weight increase and, thus, may consolidate the formation of a polychronous group. 
%----------------------------%
%

@@

\begin{figure}[H]%[t!][]
%  \centering
  \includegraphics[width=0.490\linewidth]{figures/THC_1a_k.pdf}
  \includegraphics[width=0.490\linewidth]{figures/THC_1c.pdf}
  \includegraphics[width=0.490\linewidth]{figures/THC_1b.pdf}
  \includegraphics[width=0.490\linewidth]{figures/THC_1a.pdf}
    \caption{
      Detecting event-based motifs using spiking neurons with heterogeneous delays.
    { (\textbf{a})}~Given a generic raster plot defined by a set of spikes occurring on specific addresses and at specific times, one may consider that this information consists of the repeated occurrence of groups of precise motifs of spikes that define ``polychronous groups'' (PGs). 
    { (\textbf{b})}~The generative model is defined by this set of motifs (here $4$ PGs) each defined by  synapses defined by weights (increasing with the radius of the black dots) with heterogeneous delay. 
    { (\textbf{c})}~Generalizing the core polychrony detection model (see Figure~\ref{fig:izhikevich}), one can define a layer of neurons that detect the identity and timing of these PGs~\citep{grimaldi_learning_2022}. % such that the addresses are the synapses of these neurons (which we therefore call here synapses).
    %Note that each afferent may be connected with multiple weights at different delays. The propagation of the afferent information through these delay may generate at each time step a synchronous motif on a subset of synapses. %These PG motifs were efficiently learned by a logistic regression.
    %{\bf (c)}~The output of the model provides with the predicted probability of occurrence of each PG motif at any time, which may be used to generate a spike as a Bernoulli trial, providing in this particular case with an exact identification of PGs occurrences.
    { (\textbf{d})}~Knowing the %MDPI: Please change the (a) into (d) in picture. Fixed !
 results of this detection, one may for illustration purposes highlight them by different colors in the raster plots, showing that in this synthetic example, all spikes are now associated with a PG. 
     }
  \label{fig:THC} % TODO: citer biol cyb fig:THC
\end{figure}
